---
title: 'Class 6: R Functions'
author: 'Hope (PID: A15652616)'
date: "10/14/2021"
output:
  pdf_document: default
  html_document: default
---

## A play with Rmarkdown
This is some plain text. I can make things **bold**. I can also make things *italic*.

```{r}
# This is a code chunk
plot(1:10)
```


## R functions

In today's class we are going to write a function together that grades some student work.

Question for today: 

> Q1. Write a function grade() to determine an overall grade from a vector of student homeworkassignment scores dropping the lowest single score. If a student misses a homework (i.e. has an
NA value) this can be used as a score to be potentially dropped. Your final function should be adquately explained with code comments and be able to work on an example class gradebook such as this one in CSV format: “https://tinyurl.com/gradeinput” [3pts]


```{r}
# Example input vectors to start with
student1 <- c(100, 100, 100, 100, 100, 100, 100, 90)
student2 <- c(100, NA, 90, 90, 90, 90, 97, 80)
student3 <- c(90, NA, NA, NA, NA, NA, NA, NA)
```


Let's start with student1 and find their average score.
```{r}
mean(student1)
```

But we want to drop the lowest score...
We could try the **min()** function

```{r}
min(student1)
```

We can also try **which.min()**
```{r}
which.min(student1)
```
This gives the postition of the lowest score

```{r}
# This would be the lowest score
student1[which.min(student1)]
```
To drop this value I can use minus

```{r}
student1[-which.min(student1)]
```

Let's use mean() again to get the average minus the lowest score.

```{r}
mean(student1[-which.min(student1)])
```

Start with student 2
```{r}
mean(student2[-which.min(student2)])
```
This doesn't work. It gives NA if a student is missing a homework. We need to remove the NA elements of the vector.

Lets try using something else
```{r}
mean(student2[ -which.min(student2) ], na.rm=TRUE)
```

This is not what we want. It dropped the 80 instead of the NA which is a missing homework assignment.

Lets try student3 really quick:
```{r}
mean(student3[ -which.min(student3) ], na.rm=TRUE)
```
It removed the lowest number which was 90 and it is trying to take the mean of all of the NA's and that doesn't work.

What if we make all of the NA's in the grades into a zero.
Let's try with student2. To do this we can try **is.na()**
```{r}
is.na(student2)
```
The **is.na()** function returns a logical vector where TRUE elements represent where the NA values are.

To use **is.na()** we can ask which position it is.
```{r}
which(is.na(student2))
```
To make the NA value into a zero:
```{r}
student.prime <- student2
student.prime
student.prime[which(is.na(student2)) ] = 0
student.prime
```

Now we need to put this all together to get the average score dropping the lowest score where we map NA values to zero.

```{r}
student.prime <- student2
student.prime
student.prime[which(is.na(student2)) ] = 0
student.prime
mean(student.prime[-which.min(student.prime)])

```

Lets make sure this is correct by taking out the zero on our own.
```{r}
student.prime
mean(c(100, 90, 90, 90, 90, 97, 80))
```
We got the same thing as before!

Lets try this for student 3
```{r}
student.prime <- student3
student.prime
student.prime[which(is.na(student3)) ] = 0
student.prime
mean(student.prime[-which.min(student.prime)])
```
We have a working snippet. Let's simplify by changing all of the student.prime's to x

```{r}
x <- student3
# Map NA values to 0
x[which(is.na(x)) ] = 0
# Find the mean without the lowest score
mean(x[-which.min(x)])
```

Now we can use this as the body of my function.
```{r}
grade <- function(x) {
# Make sure our scores are all numbers
  x <- as.numeric(x)
# Map NA values to 0
x[which(is.na(x)) ] = 0
# Find the mean without the lowest score
mean(x[-which.min(x)]) 
}
```
The function is made, lets try to get a grade for student 1.
```{r}
grade(student1)
```
Lets try student 2.
```{r}
grade(student2)
```

Last we can try for student3
```{r}
grade(student3)
```


Now we can read the full grade book CSV file. 
```{r}
scores <- read.csv("https://tinyurl.com/gradeinput", row.names=1)
scores
```
```{r}
grade(scores[10,])
```

Now that this works we need to apply to all scores by using the **apply()** function.

```{r}
apply(scores,1,grade)
```

To find the lowest score:
```{r}
which.min(apply(scores,1,grade))
```

> Q2. Using your grade() function and the supplied gradebook, Who is the top scoring student
overall in the gradebook? [3pts]

To find the max score:
```{r}
which.max(apply(scores,1,grade))
```

> Q3. From your analysis of the gradebook, which homework was toughest on students (i.e. obtained
the lowest scores overall? [2pts]

To find which homework was the hardest for the students:
```{r}
apply(scores,2, mean, na.rm=TRUE)
```
So hw 3 was the hardest for the students.


Adding an argument to flip dropping the lowest score on or off

```{r}
grade2 <- function(x, drop.lowest=TRUE) {
# Make sure our scores are all numbers
  x <- as.numeric(x)
# Map NA values to 0
x[which(is.na(x)) ] = 0
if(drop.lowest) {
# Find the mean without the lowest score
mean(x[-which.min(x)]) 
} else {
  mean(x)
}

} 
```

```{r}
student1 <- c(100,100,100,100,90)
grade(student1)
grade2(student1, drop.lowest=FALSE) 
```


