---
title: "class16_bimm143"
author: 'Hope (PID: A15652616)'
date: "11/18/2021"
output: html_document
---

## Background
The authors report on differential analysis of lung fibroblasts in response to loss of the developmental transcription factor HOXA1. Their results and others indicate that HOXA1 is required for lung fibroblast and HeLa cell cycle progression. In particular their analysis show that "loss of HOXA1 results in significant expression level changes in thousands of individual transcripts, along with isoform switching events in key regulators of the cell cycle". For our session we have used their Sailfish gene-level estimated counts and hence are restricted to protein-coding genes only.

# 1. Data Input

```{r}
library(DESeq2)
metaFile <- "GSE37704_metadata.csv"
countFile <- "GSE37704_featurecounts.csv"

# Import metadata and take a peak
colData = read.csv(metaFile, row.names=1)
head(colData)
```

```{r}
countData = read.csv(countFile, row.names=1)
head(countData)
```

# Q. Complete the code below to remove the troublesome first column from countData

```{r}
# Note we need to remove the odd first $length col
countData <- as.matrix(countData[,-1])
head(countData)
```


# Q. Complete the code below to filter countData to exclude genes (i.e. rows) where we have 0 read count across all samples (i.e. columns).

Tip: What will rowSums() of countData return and how could you use it in this context?

```{r}
# Filter count data where you have 0 read count across all samples.
countData = countData[-1, ]
head(countData)
```








