---
title: 'Class 12: Structural Bioinformatics II'
author: 'Hope (PID: A15652616)'
date: "11/4/2021"
output:
  pdf_document: default
  html_document: default
---

# Comparative analysis of protein structures

Using the bio3d package.

```{r}
library(bio3d)

pdb <- read.pdb("1hel")
pdb 
```

Lets use a bioinformatics method called NMA (Normal Mode Analysis) to predict the dynamics (flexibility) of this enzyme.

```{r}
modes <- nma(pdb)
plot(modes) 
```

Make a "move" of its predicted motion. We often call this a "trajectory".

```{r}
mktrj(modes, file="nma.pdb")
```

![](vmdscene.pdf)

```{r}
hits <- NULL
hits$pdb.id <- c('1AKE_A','4X8M_A','6S36_A','6RZE_A','4X8H_A','3HPR_A','1E4V_A','5EJE_A','1E4Y_A','3X2S_A','6HAP_A','6HAM_A','4K46_A','4NP6_A','3GMT_A','4PZL_A')
```

```{r}
# Download releated PDB files
files <- get.pdb(hits$pdb.id, path="pdbs", split=TRUE, gzip=TRUE)
```

Multiple structure alignment

```{r}
pdb <- pdbaln(files, fit=TRUE)
```

```{r}
pdb
```

# PCA

We will use the bio3d pca() function which is designed for protein structure data.

```{r}
# Perform PCA
pc.xray <- pca(pdb)
plot(pc.xray)
```

Make a trajectory visualization of the motion captured by the first Principal Component Analysis

```{r}
# Visualize first principal component
pc1 <- mktrj(pc.xray, pc=1, file="pc_1.pdb")
```

![](pca_1.pdf)
