---
title: "class17_bimm143"
author: 'Hope (PID: A15652616)'
date: "11/23/2021"
output:
  pdf_document: 
    latex_engine: xelatex
---

```{r}
vax <- read.csv("covid19vaccinesbyzipcode_test.csv")
head(vax)
```

# Q1. What column details the total number of people fully vaccinated? 
persons_fully_vaccinated

# Q2. What column details the Zip code tabulation area? 
zip_code_tabulation_area

# Q3. What is the earliest date in this dataset?
2021-01-05

# Q4. What is the latest date in this dataset? 
2021-11-16

We will use **lubridate** package to make life a lot easier when dealing with dates and times

```{r}
library(lubridate)
```


```{r}
today()
```

We make our 'as_of_date' column lubridate format...

```{r}
# Specify that we 
vax$as_of_date <- ymd(vax$as_of_date)
```

```{r}
today() - vax$as_of_date[1]
```

```{r}
today()- vax$as_of_date[ nrow(vax)]
```


Let's quickly look at the data structure using **skim()** function

```{r}
skimr::skim(vax)
```

# Q5. How many numeric columns are in this dataset?
9

# Q6. Note that there are “missing values” in the dataset. How many NA values there in the persons_fully_vaccinated column? 
8256 missing values

# Q7. What percent of persons_fully_vaccinated values are missing (to 2 significant figures)? 
10.17%

# Q8. [Optional]: Why might this data be missing?


# Q9. How many days between the first and last entry in the dataset? 

```{r}
vax$as_of_date[ nrow(vax)] -vax$as_of_date[1]
```
315 days between them

# Q10. How many unique dates are in the dataset (i.e. how many different dates are detailed)? 

```{r}
length( unique(vax$as_of_date))
```
46 unique dates

This makes sense because
```{r}
46*7
```

We will use **zipcodeR** package to help make sense of the zipcodes

```{r}
library(zipcodeR)
geocode_zip('92037')

```


To calculate distance between two zipcodes:
```{r}
zip_distance('92037','92109')
```


pull census data about ZIP code areas (including median household income etc.): 
```{r}
reverse_zipcode(c('92037', "92109"))
```
reverse_zipcode() pulls census data later on for any or all ZIP code areas we might be interested in

# Focus on San Diego county

```{r}
table(vax$county)
```

We will subset with base R
```{r}
inds <- vax$county == "San Diego"

head(vax[inds,])
```

Can be difficult to do this way, try **dplyr** package and its **filter()** function:
```{r}
library(dplyr)

sd <- filter(vax, county == "San Diego")

nrow(sd)
```

How many entries are there for San Diego county?
```{r}
nrow(sd)
```



# Q11. How many distinct zip codes are listed for San Diego County? 
```{r}
length(unique(sd$zip_code_tabulation_area))
```
107

# Q12. What San Diego County Zip code area has the largest 12 + Population in this dataset? 
```{r}
ind <- which.max(sd$age12_plus_population)
sd[ind,]
```
92154

# What is the population in the 92037 ZIP Code area?
```{r}
filter(sd, zip_code_tabulation_area == "92037")[1,]
```
36144 are age 5+


# Q13. What is the overall average “Percent of Population Fully Vaccinated” value for all San Diego “County” as of “2021-11-09”? 
```{r}
sd.now <- filter(sd, as_of_date == "2021-11-09")

mean(sd.now$percent_of_population_fully_vaccinated, na.rm=TRUE)
```
67.3% are fully vaccinated

```{r}
summary(sd.now$percent_of_population_fully_vaccinated)
```


# Q14. Using either ggplot or base R graphics make a summary figure that shows the distribution of Percent of Population Fully Vaccinated values as of “2021-11-09”?

Make a histogram using base R
```{r}
hist(sd.now$percent_of_population_fully_vaccinated)
```

Using ggplot instead: 
```{r}
library(ggplot2)

ggplot(sd.now) +aes(percent_of_population_fully_vaccinated) + geom_histogram(bins=15)
```

What about 92037 - UCSD/La Jolla area?
```{r}
ucsd <- filter(sd, zip_code_tabulation_area == "92037")
head(ucsd)
```

```{r}
ucsd[1,]$age5_plus_population
```
population above age 5: 36,144


# Q15. Using ggplot make a graph of the vaccination rate time course for the 92037 ZIP code area: 
```{r}
library(ggplot2)

ggplot(ucsd) +
  aes(as_of_date,
      percent_of_population_fully_vaccinated) +
  geom_point() +
  geom_line(group=1) +
  ylim(c(0,1)) +
  labs(x="Date", y="Percent Vaccinated") 
```

# Comparing 92037 to other similar sized areas
Let’s return to the full dataset and look across every zip code area with a population at least as large as that of 92037 on as_of_date “2021-11-16”. 

First subset the full 'vax' dataset to include only zipcode areas with a population as large as 92037: 
```{r}
# Subset to all CA areas with a population as large as 92037
vax.36.all <- filter(vax, age5_plus_population > 36144&
                as_of_date == "2021-11-16")

#head(vax.36.all) 
vax.36 <- filter(vax, age5_plus_population > 36144)

```

```{r}
nrow(vax.36.all)
```

How many unique zip codes have a pop. as large as 92037?
```{r}
length(unique(vax.36.all$zip_code_tabulation_area))
```




# Q16. Calculate the mean “Percent of Population Fully Vaccinated” for ZIP code areas with a population as large as 92037 (La Jolla) as_of_date “2021-11-16”. Add this as a straight horizontal line to your plot from above with the geom_hline() function?

```{r}
mean(vax.36.all$percent_of_population_fully_vaccinated, na.rm=TRUE)
```
0.66

```{r}
ggplot(ucsd) +
  aes(as_of_date,
      percent_of_population_fully_vaccinated) +
  geom_point() +
  geom_line(group=1) +
  ylim(c(0,1)) +
  labs(x="Date", y="Percent Vaccinated") + geom_hline(yintercept=0.66, col="red", linetype="dashed") 
```


# Q17. What is the 6 number summary (Min, 1st Qu., Median, Mean, 3rd Qu., and Max) of the “Percent of Population Fully Vaccinated” values for ZIP code areas with a population as large as 92037 (La Jolla) as_of_date “2021-11-16”?
```{r}
summary(vax.36.all$percent_of_population_fully_vaccinated)
```


# Q18. Using ggplot generate a histogram of this data.
```{r}
ggplot(vax.36.all) +aes(percent_of_population_fully_vaccinated) + geom_histogram(bins=15) 
```


# Q19. Is the 92109 and 92040 ZIP code areas above or below the average value you calculated for all these above?
```{r}
vax %>% filter(as_of_date == "2021-11-16") %>%  
  filter(zip_code_tabulation_area=="92040") %>%
  select(percent_of_population_fully_vaccinated)
```
52%, less than average

```{r}
vax %>% filter(as_of_date == "2021-11-16") %>%  
  filter(zip_code_tabulation_area=="92109") %>%
  select(percent_of_population_fully_vaccinated)
```
68%, above the average


# Q20. Finally make a time course plot of vaccination progress for all areas in the full dataset with a age5_plus_population > 36144.

```{r}
ggplot(vax.36) + aes(as_of_date,
      percent_of_population_fully_vaccinated,
      group=zip_code_tabulation_area) + geom_line(alpha=0.2) + geom_hline(yintercept = 0.66, col="red", linetype="dashed") + labs(x="Date", y="Percent Vaccinated")
```

# Q21. How do you feel about traveling for Thanksgiving and meeting for in-person class next Week?
It makes me nervous being around people that may have traveled far and been around big groups of people. I see how other people are not as careful as I am and it makes me uneasy.






