---
title: "class19_bimm143"
author: 'Hope (PID: A15652616)'
date: "12/5/2021"
output:
  pdf_document: default
  html_document: default
---

# Section 4: Population Scale Analysis

How many samples do we have?
```{r}
expr <- read.table("rs8067378_ENSG00000172057.6.txt")
head(expr)
```

```{r}
nrow(expr)
```

We have 462 samples in the data. 

What if we want to know the number of each type?
```{r}
table(expr$geno)
```
Now we can see there are 108 A/A, 233 A/G, and 121 G/G

Now we want to make a summary figure of the data using ggplot:
```{r}
library(ggplot2)
```


We are making a boxplot: 
```{r}
med <- aggregate(exp ~  geno, expr, median)
ggplot(expr) + aes(geno, exp, fill=geno) + geom_boxplot(notch = TRUE) + geom_text(data = med, aes(label = exp, y = exp + 2))

```

I made a table for the median values to show in the box plot. I am going to show the table:
```{r}
med
```
The medians were found above.

# Question 14: what could you infer from the relative expression value between A/A and G/G displayed in this plot? Does the SNP effect the expression of ORMDL3?
There seems to be greater expression of ORMDL3 in A/A genotypes than G/G genotypes based on the median alone but the data overall in the boxplot mostly  overlaps with each other so it's hard to say for sure. 



